Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-04-13 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Add FDS input files and coarse, medium results, update tex and bib
Adding photos and words to Chap 4, 5,6

\usepackage{graphicx}
\usepackage{rotating}
\frenchspacing

@@

@@

The gypsum wallboard used in these experiments is referred to as `regular' as opposed to gypsum wallboard `Type C' or `Type X' which are used in fire rated assemblies. The gypsum wallboard used in these experiments was purchased from same manufacturer in an effort to have the most similar panels for use between the experiments. Each 1.22~m by 2.44~m by 12.7~mm thick panel weighed approximately 30.5~kg.  The physical features of the gypsum wallboard panels are given in the table below.

@@

% first two lines of table based on measurements of GB at NIST.  Lower 4 lines from Gross 1985

@@

trim photo sets added here and compare with temps and HF

@@

The fire pattern experiments were conducted in residential scale compartments built for the purpose of these experiments.  The compartments were constructed on top of a portable wood floor platform.  Each of the test compartments consisted of three 3.6~m long by 2.4~m high wood framed walls. The wood elements of the frame were composed of kiln-dried fir with a nominal cross section of 90~mm by 38~mm.  A partial ceiling with a width of at least 1.2~m, measured from the back wall of the compartment was installed across the width of the compartment, this allowed for the plume to impact the ceiling and form a ceiling jet along the back wall, before the fire gases flowed up and out of the compartment and into the 6~m by 6~m hood in the NIST Large Fire Research Laboratory. The interior surface of the walls and ceiling was composed of 12.7~mm thick, regular core gypsum wallboard. As defined by ASTM, gypsum wallboard is designed for use on walls, ceilings, or partitions.  It is composed of a non-combustible core, essentially gypsum, with paper bonded to the surface of the core. The back wall of the compartment consisted of three full gypsum board panels, each 1.2~m wide and 2.4~m high.  The front wall of the compartment was open along the full width of the compartment to minimize the impact of entrained air being channeled through a doorway. These experiments were conducted without instrumentation in the walls for several reasons; to compare the non-instrumented gypsum wallboard fire pattern with the instrumented gypsum wallboard fire pattern results to determine if the instrumentation affected the pattern generation, to provide more realistic conditions, and to enable the conduct of additional replicate experiments given the limited laboratory availability.  Based on the outcome of the instrumented calibration target wall results, burner positions were chosen. After each burn, the fire patterns were measured, photographed and analyzed for repeatability. The fire damaged wallboard was subsequently removed and replaced with new sections of painted gypsum wallboard. The number of experiments for each case was dependent on the repeatability and similarity of the pattern development. Several types of fire pattern experiments were conducted in the compartments, as detailed in the sections to follow, to examine the impact of the insulation behind the gypsum wallboard panel on the resulting fire pattern and other series of experiments were conducted to compare the fire patterns based on the burner centered on the rear wall or in or near a corner on the rear wall of the compartment.

@@

 Impact of Wall Construction

 The initial compartment experiments were conducted to examine the impact of the method of construction of the walls on the development of the fire patterns. This data determined how the walls for the remaining compartment fire pattern experiments would be constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with fiberglas batt insulation in the voids of the wood frame representative of an exterior wall. From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occured when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels,approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

The burner was centered on the middle panel of the rear wall. The wall behind the center panel of the rear wall was insulated to met the needs of the experiment.  As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment.  Each of the painted gypsum board panels that were used for a fire pattern had a 5~cm grid drawn on it to assist with the determination of the geometry of the fire pattern.

@@

Natural gas and gasoline only

@@

@@

@@

\begin{sidewaystable}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   &   Max. Height (m)	         & Max. Width (m)               &  Height at Max. Width (m) &  Area (m$^2$)  \\ \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (12)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\
		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{sidewaystable}

>>>>

\usepackage{graphicx}
\usepackage{rotating}
\usepackage{url}
\frenchspacing

@@

Since these experiments were the culmination of the instrumented series of full scale tests, the use of gypsum wallboard was intended to provide some insight into how much fire damage would result. These experiments proved to be challenging in terms of time and effort.  Each panel of gypsum wallboard had to be carefully drilled for thermocouple and heat flux gauge placement. Since there was only one set of heat flux gauges this reduced the number experiments that could be conducted on a given day.  With other project demands, the number of instrumented wallboard experiments that could be conducted under the oxygen consumption calorimeter was reduced due to time constraints. The test matrix is shown in Table~\ref{tab:IWGB_experiments}. There was also a concern that these experiments would have limited value due to potential interference from the water-cooled heat flux gauges and potential physical damage caused by drilling the holes for them.

@@

The gypsum wallboard used in these experiments is referred to as `regular' as opposed to gypsum wallboard `Type C' or `Type X' which are used in fire rated assemblies. The gypsum wallboard used in these experiments was purchased from same manufacturer in an effort to have the most similar panels for use between the experiments. Each 1.22~m by 2.44~m by 12.7~mm thick panel weighed approximately 30.5~kg.  The thermal properties of the gypsum wallboard panels are given in the Table~\ref{tab:Gypsum wallboard_Thermal_Properties}.  The first two rows of data were measured from samples of the gypsum wallboard used in these experiments.  The bottom four rows of data are from ~cite\{Gross:1985}.

There have been a number of studies that have examined gypsum wallboard under elevated temperature conditions in terms of it's thermal properties or in terms of its physical change as a result to exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating the residue of the gypsum continues to transform and soften turning into anhydrite, commonly this process is known as calcination.  Schroder exposed samples of gyspum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X~\-ray diffraction  add Schroder.  NFPA 921 also discusses calcination and how the investigator can use the visual appearance or probing of the soft dyhydrated gypsum to map fire patterns.

@@

@@

As defined in NFPA 921, a fire is the visible or measurable physical changes, or identifiable shapes, formed by a fire effect or group of fire effects~\cite{NFPA_921}.  In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{nfpa_921}.  As presented in the introduction, DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

Figure~\ref{demarcation} is a photograph of the edge of a fire pattern generated by the natural gas burner.  It will serve as visual definition of the line of demarcation to be used in this study to determine the limits of a fire pattern.  Referring to DeHaan's classifications, there are no surface deposits such as light soot that could be brushed or blown off the gypsum wallboard.  There is a thin band of surface thermal effects in the form of discoloration of the paint, this could be considered an outer bound of the pattern. To the left of the discoloration is clearly an larger area where charring (blackening of the paint/paper), penetration (lifting of the painted/paper residue), and consumption (areas where the paint/paper reside is gone altogether).  The last two classifications or criteria will be used in this study to define the edge of the fire pattern.  Any area where the paper has been burned away or penetrated such that it has lifted away from the gypsum will be considered part of the fire pattern and define the lines of demarcation for this study.

\begin{figure}[p]
	\centering
	\includegraphics[width=\textwidth]{../Figures/demarcation}  \\
	\caption[Photograph showing clear lines of demarcation]{Photograph showing clear lines of demarcation from a fire pattern generated on painted gypsum wallboard by a natural gas fueled burner.}
	\label{demarcation}
\end{figure}

Soot and discoloration alone was not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as show in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be clearly delininated while the discolored and sooted areas tend to meld together.  Further the focus of the measurements in this study are thermal, therefore the focus is on the burned, not burned line of demarcation.

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrtaed paint/paper layer which exposed the gypsum underneath.

@@

The fire pattern experiments were conducted in residential scale compartments built for the purpose of these experiments.  The compartments were constructed on top of a portable wood floor platform.  Each of the test compartments consisted of three 3.6~m long by 2.4~m high wood framed walls. The wood elements of the frame were composed of kiln-dried fir with a nominal cross section of 90~mm by 38~mm.  A partial ceiling with a width of at least 1.2~m, measured from the back wall of the compartment was installed across the width of the compartment, this allowed for the plume to impact the ceiling and form a ceiling jet along the back wall, before the fire gases flowed up and out of the compartment and into the 6~m by 6~m hood in the NIST Large Fire Research Laboratory. The interior surface of the walls and ceiling was composed of 12.7~mm thick, regular core gypsum wallboard. As defined by ASTM, gypsum wallboard is designed for use on walls, ceilings, or partitions.  It is composed of a non-combustible core, essentially gypsum, with paper bonded to the surface of the core. The back wall of the compartment consisted of three full gypsum board panels, each 1.2~m wide and 2.4~m high.  The front wall of the compartment was open along the full width of the compartment to minimize the impact of entrained air being channeled through a doorway. These experiments were conducted without instrumentation in the walls for several reasons; to compare the non-instrumented gypsum wallboard fire pattern with the instrumented gypsum wallboard fire pattern results to determine if the instrumentation affected the pattern generation, to provide more realistic conditions, and to enable the conduct of additional replicate experiments given the limited laboratory availability. A photograph of the compartment fire experimental arrangement is shown in Figure~\ref{Compartment_Test}. In the instrumented gypsum wallboard experiments, as the burners were moved away from the position against the wall, the fire effects on the wall diminished significantly.  Therefore most of the fire pattern compartment experiments were conducted with the burners against the wall or corner of the compartment or less than one burner length from the wall.

Prior to each burn, grid lines approximately 51~mm on center were drawn on the painted gypsum wallboard panel that would be exposed to the fire.  After each burn, the fire patterns were photographed and analyzed for repeatability by looking at the portion of the grid that was burned. The fire damaged wallboard was subsequently removed, photographed again, and replaced with new sections of painted gypsum wallboard. Several types of fire pattern experiments were conducted in the compartments; to examine the impact of the wall construction on the resulting fire pattern; and to compare the fire patterns based on the burner centered on,or the rear wall, or in, or near a corner on the rear wall of the compartment.

\begin{figure}[ht!]
\includegraphics[width=\columnwidth]{../Figures/Compartment_Test} \\
	\caption[Photograph of the wall fire pattern experiments compartment arrangement]{Photograph of the wall fire pattern experiments compartment arrangement.  Experiment shown is being conducted with a polyurethane burner against the back wall.}
	\label{Compartment_Test}
\end{figure}

@@

 The initial compartment experimental series were conducted to examine the impact of the method of construction of the walls on the development of the fire patterns. These experiments would determine how the walls for the remaining compartment fire pattern experiments would be constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting,a kraft paper facing, and an "R value" of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

The burner was centered on the middle panel of the rear wall and was in contact with the wall. The wall behind the center panel of the rear wall was constructed either as open, closed, or insulated to meet the needs of the experiment. The burner was positioned against the face of the painted gypsum board. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300 second burn time which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.

@@

Ten replicate experiments were conducted with each of the three wall construction types.  The walls were exposed to both the natural gas and the gasoline burners.  For these experiments the burner was in contact with the gypsum wallboard.  A total of 60 experiments were conducted.
The results will be presented by wall construction type.  The order of presentation will be open, closed, and insulated wall construction.

@@

Figure~\ref{NG_Open_Wall} shows a comparison of the ten fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sized however they are scaled so that the images are comparable.        
The grids were used to assist with the determination of the fire pattern area and the sizing of the images. In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{nfpa_921}.

@@

\begin{sidewaystable}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   &   Max. Height (m)	         & Max. Width (m)               &  Height at Max. Width (m) &  Area (m$^2$)  \\ \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (10)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\
		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{sidewaystable}
